Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-02-04 src/article.tex
submission draft

The \anon{Tagliavini} Collection in \anon{Bologna \cite{Tagliavini2007}}, renowned for its historical keyboard instruments, exemplifies this dilemma. Housed at \anon{Museo San Colombano - Genus Bononiae}, with more than seventy early keyboard instruments, primarily early plucked stringed keyboards of Italian origin. The collection stands out as a valuable resource for musicologists, organologists and musicians alike. Preserving the instruments' authenticity was the cornerstone of \anon{Ferdinando Tagliavini}’s vision. This guiding principle led him to collect instruments which could be restored to playing condition after minimal intervention.

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—by experienced historical keyboard performers or by young musicians under strict supervision from the curator. 
To enhance accessibility and engagement, the museum commissioned the replica (Figure \ref{fig:teaser}) of a historical keyboard built in the tradition of the Italian school, which is the subject of this paper.
The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. The interface is currently connected to a commercial software sampler, though the ultimate goal is to render accessible the sound of instruments that can not be kept on a playing condition. Museum visitors are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of the interface's construction and offers reflections on its role within the \anon{Tagliavini} Collection and potential application within the wider musical instrument museum context.

\textcolor{red}{
As part of the initiative to improve the accessibility of museums, the Museo San Colombano - Tagliavini Collection, part of the cultural itinerary of Fondazione Carisbo, has designed a series of projects to improve cognitive and sensory accessibility at different levels, in a project funded by NextGenerationEU. The aim of this project is to exploit the potential of the collection of musical instruments to create an immersive experience and interaction with the heritage on display. 
reinterpretation of the historical musical objects in their tactile-auditory dimension aims to open up the exhibition to blind and partially sighted visitors. The augmented keyboard fulfils many of the project's aims and is part of the initiative to allow visitors of different abilities to engage in a hands-on creative process within the museum.
}

@@

@@

The keyboard presented here inherits some aspects from the Tromba Moderna project. The project aims to offer a tool to enhance the fruition of the \anon{Tagliavini} collection while retaining a form of continuity with historical instrument-building traditions. However, the electronics hidden inside the interface are not intended to augment or disrupt the tactile feedback; rather, they serve as the silent and invisible link between the mechanical and the digital realms. 
The optical sensing technique for the keyboard is adapted from a similar project on the piano by McPherson \cite{McPherson2013}. 
Thus, whether the interface proposed here is a \emph{new} musical interface is subject to discussion, particularly regarding the keyboard's tactile response designed to adhere to longstanding harpsichord building traditions. This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in `The \emph{O} in NIME' \cite{Masu_NIME_2023}.
The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development. This project extends the longevity of the results beyond the scope envisaged in the original works, enhancing their sustainability, as discussed by Masu \emph{et al.} \cite{Masu_NIME_2023}.

@@

\begin{itemize}
    \item \emph{Faithfulness}. The keyboard mechanism ensured fidelity to its authentic operation. The electronics system would not seek to `fix' or `improved' on limitations of the original design.     
    \item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
    \item \emph{No visible electronics}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
    \item \emph{Reduced use of space}. As museums often face space limitations, the instrument needed be compact enough not to compromise space for exhibition of the permanent display. 
\end{itemize}

These constraints demanded an electronics system that was non-invasive in design and that was respectful of the historical aesthetics of the instruments. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. 
While actuators can generate considerable force, it is difficult to achieve a jack's free motion and resistance without extensive mechanical intervention such as that carried out by Gillespie \cite{Gillespie1996}. A recreation of the mechanism can already achieve the tactile realism of the harpsichord.
It should first be determined if mechanical recreation is sufficient as part of a digital musical instrument before pursuing elaborate electromechanical emulation.

The final design, shown in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key, two-register harpsichord keyboard replica. The keyboard layout was deliberately modelled after early Italian harpsichords, leveraging the human tendency to be influenced by visual elements when making musical judgments \cite{Tsay2013}. This effect is particularly relevant in musical instruments, as demonstrated by studies conducted by Fritz \emph{et al.} \cite{Fritz2012, Fritz2014, Fritz2017}. Instead of attempting to mitigate this influence, the design embraces it. This phenomenon, akin to a `musical instrument McGurk effect’, is one of the reasons why the electronics have been concealed from view. The aesthetics of the interface enhance its likelihood of being perceived as an `authentic' musical instrument. Given the extensive research into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020}, the visual component may provide the final persuasive element necessary for acceptance, similar to the way visual perception affects judgments of musical performance \cite{Tsay2013}.

\begin{figure}[!b]  
  \centering
  \includegraphics[width=0.9\linewidth,trim={0 0 0 0.5cm},clip]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{} 
  \label{fig:jack-tags}
\end{figure}

@@

A modular system of printed circuit boards (PCBs) was designed to manage the sensors and process their output. The system uses 49 optical sensors which are distributed across seven boards that contain 7 sensors each. The PCBs were secured to the underside of wrest plank and above the keys (Figure \ref{fig:49-key-bottom}), allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

@@

\subsection{Materials and Construction}
The keyboard was designed to replicate the tactile and aesthetic sensations of playing an historical Italian harpsichord. Traditional materials were used, including walnut for the wrest plank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by Venetian harpsichords, particularly the \anon{1547 Alessandro Trasuntino} harpsichord at the \anon{San Colombano Museum}. An exception was made for the short octave --- the assigning of common keys in the first octave instead of a chromatic scale --- typically found in Italian harpsichords, which was replaced by a standard octave layout to more easily accommodate the interaction with commercial sample libraries.

@@

@@

\begin{figure}
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{
    3-Key Model Harpsichord Mechanism    
    %acquired from \anon{San Colombano, Bologna}, 2023.
    }
    \label{fig:3key}
\end{figure}

@@

\begin{figure}[!h] 
  \centering
  \includegraphics[width=0.7\linewidth,trim={0 0.5cm 0 0.5cm},clip]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer.}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth,trim={0 1.5cm 0 1.5cm},clip]{src/images/sensor-board-w-baffles.jpeg}
    \\
    \includegraphics[width=0.8\linewidth,trim={0 2cm 0 2.5cm},clip]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}

@@

\begin{figure}[!b] 
  \centering
  \includegraphics[width=\linewidth]{src/images/sensor-board-reverse-side-anon.png} 
  \caption{Reverse side of the sensor board showing the RGB LEDs and terminal connections. The \texttt{SIG} terminal (Top Left) is routed to an ADC channel. Power \texttt{5V, 3V, GND}, multiplexer channels \texttt{A, B, C} and LED PWM input \texttt{DI} and output \texttt{DO}  connect from the previous sensor board (Left) and are routed to the next (Right).}
  \Description{} 
  \label{fig:sensor-reverse}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.8\linewidth,trim={0 5cm 0 0},clip]{src/images/mac-mini.jpg}
    \caption{Components and cabling hidden in the bottom section of the instrument case.}
    \label{fig:mac-mini}
\end{figure}

@@

For the initial version of the exhibition, a harpsichord sample library was used and controlled via Native Instrument's Kontakt\footnote{\href{https://www.native-instruments.com/en/products/komplete/samplers/kontakt-8/?srsltid=AfmBOorKUf43SoIxGBS2-GnXmKHHkgcfcfWRskpweDhLSG3FiF0qrf2w}{Kontakt webpage} (accessed 31 Jan 2025)}. The software was installed on a Mac Mini hosted inside the instrument's case. Holes were drilled at the back of the instrument's frame, away from sight, to allow the passage of cables, such as from power supplies, USB connectors and headphone jack (Figure \ref{fig:mac-mini}). 
An iPad is used as a monitor through which visitors can adjust playback parameters such as tuning, voicing and which stops are engaged. The next iteration will use a library of samples from instruments in the collection, which the museum is currently carrying out.

@@

The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of 20 people -- consisting of staff, surveillance personnel and visitors who were given training on the system -- and expert feedback from \anon{Catalina Vicens} and \anon{Roberto Livi}.
The collection of visitor feedback was carried out as short informal interviews and primarily as a means of identifying any technical problems.
The keyboard shows promise in enhancing the museum experience as comments about a ``sense of disconnect'' found in the Benton Fletcher Collection \cite{McAlpine2014} were absent in this initial feedback.

@@

>>>>

The \anon{Tagliavini} Collection in \anon{Bologna \cite{Tagliavini2007}}, housed at \anon{Museo San Colombano - Genus Bononiae} and renowned for its historical keyboard instruments, exemplifies this dilemma. With over fifty early keyboard instruments, primarily early plucked stringed keyboards of Italian origin, the collection stands out as a valuable resource for musicologists, organologists and musicians alike. Preserving the instruments' authenticity was the cornerstone of \anon{Ferdinando Tagliavini}’s vision. This guiding principle led him to collect instruments that could be restored to their playing condition after minimal intervention.

However, the delicate mechanisms of these instruments and their historical significance mean they are only played under strict conditions—by experienced historical keyboard performers or young musicians under strict supervision from the curator. 
To enhance accessibility and engagement, the museum commissioned the replica (Figure \ref{fig:teaser}) of a historical keyboard built in the tradition of the Italian school, which is the subject of this paper.
The keyboard, which museum visitors can use without special permission, presents two choirs of muted strings and can generate MIDI messages. The interface is presently linked to a commercial software sampler; however, the ultimate aim is to make the sound of instruments in the collection that can no longer be maintained in playable condition accessible. Museum visitors are invited to play the interface and listen through a pair of headphones. This work outlines the technological aspects of the interface's construction and offers reflections on its role within the \anon{Tagliavini} Collection and potential application within the wider musical instrument museum context.

@@

As part of a broader initiative to enhance accessibility in museums, \anon{the Museo San Colombano - Tagliavini Collection}, within the cultural itinerary of \anon{Fondazione Carisbo}, has developed a series of projects aimed at improving cognitive and sensory access, supported by funding from NextGenerationEU. The project seeks to leverage the museum’s collection of musical instruments to create an immersive and interactive engagement with the exhibited heritage. The reinterpretation of historical musical objects through their tactile and auditory dimensions is intended to make the exhibition more accessible to blind and partially sighted visitors. The augmented keyboard aligns with these objectives, serving as an important component of the initiative to provide visitors of varying abilities with a hands-on, creative experience within the museum.

@@

The keyboard presented here inherits some aspects from the Tromba Moderna project. The project aims to offer a tool to enhance the visitor experience of the \anon{Tagliavini} collection while retaining a form of continuity with historical instrument-building traditions. However, the electronics hidden inside the interface are not intended to augment or disrupt the tactile feedback; rather, they serve as the silent and invisible link between the mechanical and the digital realms.

The optical sensing technique for the keyboard is adapted from a similar project on the piano by McPherson \cite{McPherson2013}. As such, this project extends the longevity of the results beyond the scope envisaged in the original works, enhancing their sustainability, as discussed by Masu \emph{et al.} \cite{Masu_NIME_2023}. The intended use of the keyboard through meaningful interaction with a museum exhibit is where the novelty of this work lies, rather than solely in its technological development.

@@

\begin{itemize}
    \item \emph{Faithfulness}. The keyboard mechanism had to ensure fidelity to its authentic operation. The electronics system would not seek to `fix' or `improve' the limitations of the original design.     
    \item \emph{Robustness and reliability}. The system needed to accommodate frequent use by museum visitors and allow for straightforward maintenance by staff without requiring specialised technical expertise.
    \item \emph{No visible electronic components}. To preserve the visual integrity of the exhibit, all electronic components had to remain concealed except for a pair of headphones and a small display for audio parameter adjustments.
    \item \emph{Reduced use of space}. As museums often face space limitations, the instrument needed be compact enough not to compromise space for exhibition of the permanent display. 
\end{itemize}

These constraints demanded an electronics system that was non-invasive in design and respectful of the instruments' historical aesthetics. The requirement for faithfulness to the original mechanism immediately ruled out a mechanical design reliant on electromechanical actuators, as in previous works on piano haptics \cite{Timmermans2020,Gillespie1996}. 
While actuators can generate considerable force, achieving a jack's free motion and resistance is difficult without extensive mechanical intervention such as that carried out by Gillespie \cite{Gillespie1996}. 
% Recreating the mechanism can already achieve the tactile realism of the harpsichord.
% It should first be determined if mechanical recreation is sufficient as part of a digital musical instrument before pursuing elaborate electromechanical emulation.

The final design, shown in Figures \ref{fig:teaser} and \ref{fig:details}, is a 49-key, two-register harpsichord keyboard replica. Since the Tagliavini collection hosts primarily early plucked stringed keyboards of Italian origin, the keyboard layout was deliberately modelled after early Italian harpsichords, leveraging the human tendency to be influenced by visual elements when making musical judgments \cite{Tsay2013}. This effect is particularly relevant in musical instruments, as demonstrated by studies conducted by Fritz \emph{et al.} \cite{Fritz2012, Fritz2014, Fritz2017}. Instead of attempting to mitigate this influence, the design embraces it. This phenomenon, akin to a `musical instrument McGurk effect’, is one of the reasons why the electronics have been concealed from view. The aesthetics of the interface enhance its likelihood of being perceived as an `authentic' musical instrument. Given the extensive research into modelling and recreating piano action \cite{Cadoz1990, Gillespie1996, Timmermans2020}, the visual component may provide the final persuasive element necessary for acceptance, similar to the way visual perception affects judgments of musical performance \cite{Tsay2013}.

\begin{figure}[!b]  
  \centering
  \includegraphics[width=\linewidth,trim={0 0 0 0.5cm},clip]{src/images/tagging-jacks-3.jpg} 
  \caption{Gradient stickers applied to the side of the jack body. The coarse gradient scale was selected to maximise signal excursion while preserving signal readout stability.}
  \Description{} 
  \label{fig:jack-tags}
\end{figure}

@@

A modular system of printed circuit boards (PCBs) was designed to manage the sensors and process their output. The system uses 49 optical sensors distributed across seven boards containing 7 sensors each. The PCBs were secured to the underside of the wrest plank and above the keys (Figure \ref{fig:49-key-bottom}), allowing them to be adjusted during installation. Ribbon cables connected the PCBs, providing flexibility during assembly while maintaining a compact form factor. Additional modifications, including baffles and adhesive improvements, were made to optimise the reliability of the sensor system during calibration and use.

@@

\subsection{Materials and Construction}
The keyboard was designed to replicate the tactile and aesthetic sensations of playing a historical Italian harpsichord. Traditional materials were used, including walnut for the wrest plank, chestnut for the key levers, boxwood and ebony for key covers, and cypress for the case and soundboard. The 98 jacks were made from beech, fitted with brass springs and natural seagull feather plectra. The design was inspired by the \anon{1547 Alessandro Trasuntino} harpsichord at the \anon{San Colombano Museum}, considered one of the first prototypes to have been conceived with two sets of 8-foot strings. An exception was made for the short octave --- the assigning of common keys in the first octave instead of a chromatic scale --- typically found in Italian harpsichords, which was replaced by a standard octave layout to more easily accommodate the interaction with commercial sample libraries.

@@

\begin{figure}[!t]
    \centering    
    \includegraphics[width=\linewidth]{src/images/3-key-side.png}
    \caption{
    3-Key Model Harpsichord Mechanism    
    %acquired from \anon{San Colombano, Bologna}, 2023.
    }
    \label{fig:3key}
\end{figure}

@@

@@

\begin{figure}[!b] 
  \centering
  \includegraphics[width=0.7\linewidth,trim={0 0.5cm 0 0.5cm},clip]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer.}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.8\linewidth,trim={0 2cm 0 2.5cm},clip]{src/images/baffles.png}
    \caption{Baffles designed to prevent cross-talk between adjacent sensors.}
    \Description{}
    \label{fig:baffles}
\end{figure}

@@

@@

@@

\begin{figure}[!b] 
  \centering
  \includegraphics[width=\linewidth,trim={0 1.5cm 0 1.5cm},clip]{src/images/sensor-board-w-baffles.jpeg}
    \\
  \includegraphics[width=\linewidth]{src/images/sensor-board-reverse-side-anon.png} 
  \caption{Sensor board showing the sensor side (top) and rear side with RGB LEDs and terminal connections (bottom). The \texttt{SIG} terminal (Top Left) is routed to an ADC channel. Power \texttt{5V, 3V, GND}, multiplexer channels \texttt{A, B, C} and LED PWM input \texttt{DI} and output \texttt{DO}  connect from the previous sensor board (Left) and are routed to the next (Right).}
  \Description{} 
  \label{fig:sensor-reverse}
\end{figure}

For the initial version of the exhibition, a harpsichord sample library was used and controlled via Native Instrument's Kontakt\footnote{\href{https://www.native-instruments.com/en/products/komplete/samplers/kontakt-8/?srsltid=AfmBOorKUf43SoIxGBS2-GnXmKHHkgcfcfWRskpweDhLSG3FiF0qrf2w}{Kontakt webpage} (accessed 31 Jan 2025)}. The software was installed on a Mac Mini hosted inside the instrument's case. Holes were drilled at the back of the instrument's frame, away from sight, to allow the passage of cables, such as from power supplies, USB connectors and headphone jack (Figure \ref{fig:mac-mini}). 
An iPad is used as a monitor through which visitors can adjust playback parameters such as tuning, voicing and which stops are engaged.

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim={0 5cm 0 0},clip]{src/images/mac-mini.jpg}
    \caption{Components and cabling hidden in the bottom section of the instrument case.}
    \label{fig:mac-mini}
\end{figure}

@@

The exhibition is set to open approximately a month after this work's submission, and only preliminary feedback has been collected from a pool of 20 people -- consisting of staff, surveillance personnel and visitors who were given training on the system -- and expert feedback from \anon{Catalina Vicens} and \anon{Roberto Livi}.
Visitor feedback was collected through short informal interviews and primarily to identify technical problems.
The keyboard shows promise in enhancing the museum experience as comments about a ``sense of disconnect'' found in the Benton Fletcher Collection \cite{McAlpine2014} were absent in this initial feedback.

@@

This work has incorporated and developed further the ideas of reusing and complimenting the `old' in digital musical instrument design as put forward in the article `The \emph{O} in NIME' \cite{Masu_NIME_2023}. By integrating historical keyboard-building traditions with digital augmentation, it offers a possibility to extend the practical lifespan of musical heritage while maintaining the tactile qualities of the original instrument. Rather than prioritising technical novelty, the project demonstrates how digital interventions can support meaningful interaction with a shared musical heritage, ensuring its continued relevance in contemporary museum contexts.

### 2025-02-04 src/main.tex
submission draft

%%
%% This is file `sample-nime-paper.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-nime-paper.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% History of NIME Template Updaters:
%% Modified by Florent Berthaut, Charles Martin, Yichen Wang, 21 November 2024
%% Modified by Adnan Marquez-Borbon 30 November 2022
%% Modified by Courtney Reed 28 November 2022
%% Modified by Joe Wright 14 December 2019
%% Modified by Niccolò Granieri 10 October 2018
%% Modified by Angelo Fraietta 23 December 2018
%% Modified by Angelo Fraietta 22 November 2018
%% Modified by Rodrigo Schramm on 22 September 2018
%% Modified by Luke Dahl on 17 October 2-17
%% Modified by Cumhur Erkut on <2016-10-11 Tue>
%% Modified by Edgar Berdahl on 5 November 2014
%% Modified by Baptiste Caramiaux on 25 November 2013
%% Modified by Kyogu Lee on 7 October 2012
%% Modified by Georg Essl on 7 November 2011
%%
%% And here's the template:
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{nimeart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{nimeart} or whichever template is required
%% for your publication.
%%
%%
% \documentclass[sigconf]{nimeart}
\documentclass[sigconf,anonymous]{nimeart}
% \documentclass[sigconf,anonymous,review]{nimeart}
\usepackage{graphicx,subcaption,svg}

@@

\author{Roberto Livi}
\affiliation{%
  \institution{Master Harpsichord Maker \\ Museo San Colombano}
  \city{Pesaro}
  \country{Italy}
  }
\email{robertvs.livi@gmail.com}

@@

This paper describes the design and creation of an electronically augmented replica of a harpsichord keyboard with a typical early Italian layout to create a digital musical instrument. 
The keyboard was commissioned for exhibition in a musical instrument museum to enhance the visitor experience by providing an interface to digitised versions of instruments within the collection. 
The replica attempts to balance the competing demands of historical authenticity, public accessibility, and preservation. 
It replicates the original instrument's tactile feedback and mechanical resistance by employing historically informed construction techniques. Optical sensors integrated within the mechanism capture the jacks' motion data, enabling MIDI message generation. 
This work situates itself within broader discussions on the role of technology in museums. 
A keyboard interface of this type offers an opportunity to enhance visitor interaction with cultural heritage while safeguarding delicate artefacts. 
The paper examines the keyboard's design principles, technical implementation, and implications, emphasising its contribution to public engagement and the long-term preservation of musical heritage.

@@

\begin{acks}
This work received funding from the European Commission's Horizon2020 programme, through grant \anon{ERC-StG 950084 NEMUS}

>>>>

%%
%% This is file `sample-nime-paper.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-nime-paper.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% History of NIME Template Updaters:
%% Modified by Florent Berthaut, Charles Martin, Yichen Wang, 21 November 2024
%% Modified by Adnan Marquez-Borbon 30 November 2022
%% Modified by Courtney Reed 28 November 2022
%% Modified by Joe Wright 14 December 2019
%% Modified by Niccolò Granieri 10 October 2018
%% Modified by Angelo Fraietta 23 December 2018
%% Modified by Angelo Fraietta 22 November 2018
%% Modified by Rodrigo Schramm on 22 September 2018
%% Modified by Luke Dahl on 17 October 2-17
%% Modified by Cumhur Erkut on <2016-10-11 Tue>
%% Modified by Edgar Berdahl on 5 November 2014
%% Modified by Baptiste Caramiaux on 25 November 2013
%% Modified by Kyogu Lee on 7 October 2012
%% Modified by Georg Essl on 7 November 2011
%%
%% And here's the template:
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{nimeart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{nimeart} or whichever template is required
%% for your publication.
%%
%%
% \documentclass[sigconf]{nimeart}
\documentclass[sigconf]{nimeart}
% \documentclass[sigconf,anonymous,review]{nimeart}
\usepackage{graphicx,subcaption,svg}

@@

\author{Roberto Livi}
\affiliation{%
  \institution{Master Harpsichord Maker}
  \city{Pesaro}
  \country{Italy}
  }
\email{robertvs.livi@gmail.com}

@@

This paper describes the design and creation of an electronically augmented replica of a historical harpsichord keyboard with a typical 17th-century Italian layout to create a digital musical instrument. 
The keyboard was commissioned for exhibition in a musical instrument museum to enhance the visitor experience by providing an interface to digitised versions of instruments within the collection. 
The replica balances the competing demands of historical authenticity, public accessibility, and preservation. 
It replicates the original instrument's tactile feedback and mechanical resistance using historically informed construction techniques. Optical sensors integrated within the mechanism capture the jacks' motion data, enabling MIDI message generation. 
This work situates itself within broader discussions on the role of technology in museums. 
A keyboard interface of this type offers an opportunity to enhance visitor interaction with musical heritage while safeguarding delicate artefacts. 
The paper examines the keyboard's design principles, technical implementation, and implications, emphasising its contribution to public engagement and the long-term preservation of musical heritage.

@@

\begin{acks}
The construction of the keyboard received funding from NextGenerationEU within the framework of the accessibility project for the Removal of Physical and Cognitive Barriers in Museums, Libraries and Archives: Sound Heritage No Limits, in the Museum of San Colombano in Bologna - CUP: C38J23000040001 - in the context of PNRR, Mission 1. MH and MD's work was supported by the European Commission's Horizon2020 programme through grant \anon{ERC-StG 950084 NEMUS}.
